Processing large surveys with package MRP-KIT

### 2021-07-02 case-study/Munging_CCES.Rmd
checkpointing

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, message=FALSE, error=FALSE, warning=FALSE, comment=NA, out.width='95%', collapse=TRUE, prompt=TRUE)
```

<style type="text/css">
.table { width: 85%; }
div.quote-container blockquote {  font-size: small; }
</style>

Multi-level Regresssion and Postrativication (MRP)
is a two-step process for correcting model estimates
obtained from a survey sample when the survey population is not
the same as the target population over which we wish to do inference.
The first step is the doing a multilevel regression on the survey sample;
the second step is poststratification to the target population.
Conceptually, MRP is relatively simple.
In practice, it requires more than twice as much data processing.
In addition to managing the survey data, you must also obtain
and manage the target population data, in addition you must
develop a mapping between information in these two datasets.
In order to produce verifiable, replicable results, it is
critical that this mapping be documented and scripted.

@@

In this case study, the survey sample is one year's worth of data
from is Cooperative Congressional Election Study (CCES),
a bi-annual survey of of 60,000 American adults from
all 50 states and the District of Columbia.
The CCES is designed to capture 
the views and experiences of Amercians with respect to Congress
and congressional (and presidential) elections.
While this survey sample is representative at the national level
it is not large enough to provide a representative sample for sub-populations
of interest.
The population sample is the 2018 American Community Survey (ACS), the
annual micro-census conducted by the US Census Bureau.  It is designed
to be a representative sample of the US population as a whole; data is
collected from 3 million households.

Big data is precious data, but a researcher's time is even more precious.
A secondary goal of this case study is showing how to work effectively
and efficiently with large samples to produce replicable results.

### Survey sample: the 2018 CCES Common Content Dataset

The 2018 CCES Common Content Dataset is 
available from the Harvard Dataverse project:
https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi%3A10.7910/DVN/ZSBZ7K
The raw survey data is distributed in several formats from the Harvard Dataverse
as  [cces18_common_vv](https://dataverse.harvard.edu/file.xhtml?persistentId=doi:10.7910/DVN/ZSBZ7K/C1YVSF&version=6.0).
While the tabular versions of the data file are smaller, the Stata version `cces18_common_vv.dta`
contains metadata on all the questions: the question name, label, description, response labels and values.
The corresponding 
[2018 CCES Guide](https://dataverse.harvard.edu/file.xhtml?persistentId=doi:10.7910/DVN/ZSBZ7K/WZWCZ1&version=6.0)
provides a human-readable version of this metadata as well.

@@

We download the Stata file because we will need the metatdata to map the CCES sample to the ACS using MRP-KIT,
then use the [`readstata13`](https://sjewo.github.io/readstata13/) library function
[`read.dta13`](https://sjewo.github.io/readstata13/reference/read.dta13.html) to load the Stata file into R.
The result is an R data.frame with attributes containing the metadata.
```{r}
library(readstata13)
cces_2018 = read.dta13("cces18_common_vv.dta", convert.factors = T, nonint.factors = T, generate.factors=T)
```

Fist we check the type, shape and contents of the sample.
```{r}
str(cces_2018, max.level=2, list.len=12, give.attr=FALSE)
dim(cces_2018)
names(cces_2018)[1:12]
```

The Common Content portion of the survey consists of questions asked of all subjects.
Profile questions provide demographic information,
and opinion questions provide outcomes of interest.
The demographic variables include
gender, education level, race, and geographic location, among others.
```{r}
table(cces_2018$gender[,drop=TRUE])
table(cces_2018$educ[,drop=TRUE])
table(cces_2018$race[,drop=TRUE])
```
The political opinion questions on the survey are labelled "Abortion", "Immigration", "Taxes", "Health Care".
```{r}
attr(cces_2018, "var.labels")[which(startsWith(attr(cces_2018, "var.labels"), "Health Care"))[1]]
health_care_1 <- which(startsWith(attr(cces_2018, "var.labels"), "Health Care"))[1]
table(cces_2018[[health_care_1]][, drop=TRUE])
```

Of the 525 columns in the survey, we are interested in the set of political opinion questons
and a subset of the demographics.  Although not strictly necessary, we create a new dataframe
which contains just these columns of interest.
The survey metadata includes column names, and question and response labels.
We use the column names to select demographics and column labels to select political opinion questions.
```{r}
col_labels <- attr(cces_2018, "var.labels")
col_names <- attr(cces_2018, "names")
demo_cols <- c(1, which(col_names == "birthyr"),
              which(col_names == "gender" ),
              which(col_names == "educ"),
              which(col_names == "race"),
              which(col_names == "hispanic"),
              which(col_names == "marstat"),
              which(col_names == "inputstate"))
outcome_cols <- c(which(startsWith(col_labels, "Abortion --")),
                 which(startsWith(col_labels, "Immigration --")),
                 which(startsWith(col_labels, "Taxes --")),
                 which(startsWith(col_labels, "Health Care --")),
                 which(startsWith(col_labels, "President 20")))
col_labels[demo_cols]
col_labels[outcome_cols]
keep_cols <- sort(c(demo_cols, outcome_cols))
cces_2018_subset <- subset(cces_2018[keep_cols])
```

@@

The CCES data is roughly representative of the state-level political sentiment as a whole,
but if we seek to understand certain subsets of the population,
the sample may well be unbalanced.
There are only 60,000 survey responses,
but there are over 2 million
possible combinations of gender, education,
race, marital status, birthyear, and state of residence.
If we bin ages by decade, there are a mere 200,000 combinations,
binning states by region further reduces this to 16,000, etc., etc.
However, if we wish to consider state-level subpopulations of interest,
the survey data is likely to be sparse and not representative of the general population.
To address the former we do multi-level modeling; for the latter, poststratification.

### Population sample:  the 2018 American Community Survey

The US Census data is the largest representative population sample that we have access to.
A full census is conducted every 10 years.
The American Community Survey (ACS) is a micro-census conducted every year by the Census Bureau
over 1% of the households in the US, i.e., it has 3 million respondants (as opposed to 300 million in the full census).

@@

The `ipumsr` package uses several [tidyverse](https://www.tidyverse.org/) packages
including [`dplyr`](https://dplyr.tidyverse.org/) and [`haven`](https://haven.tidyverse.org/).
The [`read_ipums_micro` function](http://tech.popdata.org/ipumsr/reference/read_ipums_micro.html)
returns a single [`tbl_df`](https://tibble.tidyverse.org/reference/tbl_df-class.html) data frame, i.e., a "tibble".
```{r}
library(ipumsr)
load("acs_2018_subset.RData")
str(acs_2018_subset, max.level=2, list.len=6)
dim(acs_2018_subset)
names(acs_2018_subset)
```
We selected variables "RACE" and "HISPAN"; the extract contains two columns for each:
"RACE" and "RACED", and "HISPAN" and "HISPAND".  The "D" columns provide an extremely fine-grained
set of categories for race and hispanic origin.  As these go far beyond the categories from
the CCES which are provided, we ignore them.

The columns of this tibble are  [labelled](https://haven.tidyverse.org/reference/labelled.html) vectors.
Labelled vectors can have text labels associated with specific values.
To see how this works, we compare the values and labels for `AGE`.
The 2018 ACS data spans ages 0 through 96,
of these, only ages 0 and 90 have labels.
The function [`ipums_val_labels`](https://tech.popdata.org/ipumsr/reference/ipums_var_info.html)
allows us to inspect these labels.
```{r}
length(table(acs_2018_subset$AGE))
ipums_val_labels(acs_2018_subset$AGE)
```

The function [`as_factor`](https://haven.tidyverse.org/reference/as_factor.html)
allows us to treat labelled vectors as factors.
See the vignette https://tech.popdata.org/ipumsr/reference/ipums_var_info.html for more details.

### Aligning the Samples with MRP-KIT

We use MRP-KIT to map demographics of interest on a column-by-column basis.^[
If information is spread across multiple columns in either the survey sample
or the target population sample, it must be combined into a single column's
worth of information as MRP-KIT doesn't provide functions to do this.
*Note: all pre-processing should be documented, scripted, and **tested**!*]

In the following sections we examine each variable in turn.
This is a tedious process, therefore, we provide the following
summary of the demographic construct and the kinds of challenge each mapping presents

Construct | CCES | ACS | Challenge
--------- | ---- | --- | ---------
Sex | gender | SEX | (none)
State | inputstate | STATEFIP | specify 1-1 mapping programatically
Marital status | marstat | MARST | choice of mapping
Education | educ | EDUC | choice of mapping
Race/Hispanic ethnicity | race, hispan | RACE, HISPAN | cannot align CCES race to ACS RACE
Age | birthyr | AGE | different units, too fine-grained

#### Trivial mapping: respondant sex, 2 categories

The CCES and ACS samples record the sex of the respondant;
in the CCES this is named "gender" and in the ACS this is named "SEX".
In both, the response labels are "Male", "Female".
Creating a mapping between the two is straightforward:
we create a new SurveyQuestion object with name "sex",
then create a mappng from the CCES to the ACS,
both for column names and response labels
where for both mappings, the first element
is the survey sample name/label, and the second element
is the population sample name/label.

@@

We examine the state names from both the CCES and ACS; the same labels are used in both surveys.
```{r}
# CCES data is a factor, levels are the labels
inputstate <- levels(cces_2018_subset$inputstate[,drop=TRUE])

@@

We need to create a SurveyQuestion object, as above,  but it would be
extremely tedious to write out a pairlist, e.g.:
```
values_map=list("Alabama" = "Alabama", ..., "Wyoming" = "Wyoming")
```
[R Pairlists](https://github.com/hadley/r-internals/blob/master/pairlists.md) are
a kind of linked list data structure where each element in the list has a name.
In order to create a pairlist programmatically, we use the R function `setNames`,
which takes as its first argument an object to be labelled
and as its second argument the corresponding character vector of labels,
i.e., the expression `as.list(c("value1", "value2"),  c("tag1", "tag2")`
is equivalent to the expression `list("tag1" = "value1", "tag2" = "value2").
```{r}
mapping = as.list(setNames(STATEFIP, inputstate))
mapping[1:3]
```

Having established the values map, we create the corresponding SurveyQuestion object.
We use accessor methods `name()`, `col_names()` and `values()` to check our work.
```{r}
q_state <- SurveyQuestion$new(name = "state",
	col_names = c("inputstate", "STATEFIP"),
	values_map= mapping)
q_state$name()
q_state$col_names()
head(q_state$values())
```

#### Choice of mapping: education, marital status

The CCES and ACS samples record the education level of the respondant
in variables `educ` and `EDUC`, respectively.
Both the CCES and ACS record the highest level of education completed,
but the CCES records the degree obtained whereas the ACS records the year completed.
```{r}
names(table(cces_2018_subset$educ))
ipums_val_labels(acs_2018_subset$EDUC)
```
We choose to align these categories as follows:

CCES | ACS
---- | ---
"No HS" | "Nursery school to grade 4"
"No HS" | "Grade 5, 6, 7, or 8"
"No HS" | "Grade 9"
"No HS" | "Grade 10"
"No HS" | "Grade 11"
"High School Graduate" | "Grade 12"
"Some College" | "1 year of college"
"2-year" | "2 years of college"
"2-year" | "3 years of college"
"4-year" | "4 years of college"
"Post-grad" | "5+ years of college"
"skipped" | "N/A or no schooling"
"not asked" | "N/A or no schooling"

Alternative mappings are equally viable.
With MRP-KIT, the choice of mappings is recorded and
easily accessible.
```{r}
educ_mapping = list(
    "No HS" = "Nursery school to grade 4",
    "No HS" = "Grade 5, 6, 7, or 8",
    "No HS" = "Grade 9",
    "No HS" = "Grade 10",
    "No HS" = "Grade 11",
    "High School Graduate" = "Grade 12",
    "Some College" = "1 year of college",
    "2-year" = "2 years of college",
    "2-year" = "3 years of college",
    "4-year" = "4 years of college",
    "Post-grad" = "5+ years of college",
    "skipped" = "N/A or no schooling",
    "not asked" = "N/A or no schooling"
)
q_educ <- SurveyQuestion$new(name = "educ",
    col_names = c("educ", "EDUC"),
    values_map= educ_mapping)
q_educ$name()
q_educ$col_names()
head(q_educ$values())
```

As with eduction, the CCES and ACS samples record the marital status
of the respondant differently and we have to use our judgement as to how best
to align the two.
```{r}
names(table(cces_2018_subset$marstat))
ipums_val_labels(acs_2018_subset$MARST)
```

CCES | ACS
---- | ---
"Married" | "Married, spouse present"
"Married" | "Married, spouse absent"
"Separated" | "Separated"
"Divorced" | "Divorced"
"Widowed" | "Widowed"
"Never married" | "Never married"
"Domestic / civil partnership" | "Never married"
"skipped" | "Never married"
"not asked" | "Never married"

We note that for the 2018 CCES survey, the marital status of all respondants is known; nonetheless,
we provide a complete mapping.

```{r}
marstat_mapping = list(
    "Married" = "Married, spouse present",
    "Married" = "Married, spouse absent",
    "Separated" = "Separated",
    "Divorced" = "Divorced",
    "Widowed" = "Widowed",
    "Never married" = "Never married",
    "Domestic / civil partnership" = "Never married",
    "skipped" = "Never married",
    "not asked" = "Never married"
)
q_marstat <- SurveyQuestion$new(name = "marstat",
    col_names = c("marstat", "MARST"),
    values_map= marstat_mapping)
q_marstat$name()
q_marstat$col_names()
head(q_marstat$values())
```

#### Information split across survey columns:  race and ethnicity.

We ignore the politics of race and racial categories
and instead focus on how best to map the CCES information to the ACS information.
Both surveys have multiple variables which provide information about race and ethnicity,
and both record race and hispanic origin separately.
However the CCES treats hispanic origin as race as well as ethnic origin; the ACS does not.
```{r}
names(table(cces_2018_subset$race))
names(table(cces_2018_subset$hispan))
ipums_val_labels(acs_2018_subset$RACE)
ipums_val_labels(acs_2018_subset$HISPAN)
```
In this case, munging the raw survey data is unavoidable.
For each survey we create a new variable which blends race and hispanic ethnicity.
We are scripting everyting, so we add this to our processing scripts.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, message=FALSE, error=FALSE, warning=FALSE, comment=NA, out.width='95%', collapse=TRUE, prompt=TRUE)
options(digits=2)
if (!require("ipumsr")) stop("This case study requires ipumsr package. It can be installed using the following command: install.packages('ipumsr')")
if (!require("labelled")) stop("This case study requires labelled package. It can be installed using the following command: install.packages('labelled')")
library(ipumsr)
library(labelled)
```

<style type="text/css">
.table { width: 85%; }
div.quote-container blockquote {  font-size: small; }
blockquote {  font-size: small; }
</style>

Multi-level Regresssion and Postrativication (MRP)
is a two-step process for correcting model estimates
obtained from a survey sample when the survey population is not
the same as the target population over which we wish to do inference.
The first step is the doing a multilevel regression on the survey sample;
the second step is poststratification to the target population.
Conceptually, MRP is relatively simple.
In practice, it requires more than twice as much data processing.
In addition to managing the survey data, you must also obtain
and manage the target population data.
Most importantly, you must develop a mapping between information
in these two datasets.
In order to produce verifiable, replicable results, it is
critical that this mapping be documented and scripted.

@@

In this case study, the survey data comes from
the 2018 Cooperative Congressional Election Study (CCES),
a bi-annual survey of of 60,000 American adults from
all 50 states and the District of Columbia.
The CCES is designed to capture 
the views and experiences of Amercians with respect to Congress
and congressional (and presidential) elections.
The population data comes from the 2018 American Community Survey (ACS),
the annual micro-census conducted by the US Census Bureau.  It is designed
to be a representative sample of the US population as a whole; data is
collected from 3 million households.

Big data is precious data, but a researcher's time is even more precious.
A secondary goal of this case study is showing how to work effectively
and efficiently with large surveys to produce replicable results.

### Survey data: the 2018 CCES Common Content Dataset

The zero-th step in any data analysis is obtaining the data and checking its contents.
When working with your own data, you have already done this.
When working with someone else's data, you need to understand
both the contents of the dataset and its structure.

The 2018 CCES Common Content Dataset is 
available from the Harvard Dataverse project:
https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi%3A10.7910/DVN/ZSBZ7K
The raw survey data is distributed in several formats from the Harvard Dataverse
as  [cces18_common_vv](https://dataverse.harvard.edu/file.xhtml?persistentId=doi:10.7910/DVN/ZSBZ7K/C1YVSF&version=6.0).
While the tabular versions of the data file are smaller, the Stata version `cces18_common_vv.dta`
contains metadata on all the questions: the question name, label, description, response labels and values.
The corresponding 
[2018 CCES Guide](https://sda.berkeley.edu/sdaweb/docs/cces2018/DOC/CCES+Guide+2018.pdf)
provides a human-readable version of this metadata as well.

@@

We download the Stata file because we will need the metatdata to map the CCES sample to the ACS using MRP-KIT,
then use the package [`readstata13`](https://www.rdocumentation.org/packages/readstata13/versions/0.10.0)
function [`read.dta13`](https://www.rdocumentation.org/packages/readstata13/versions/0.10.0/topics/read.dta13)
to load the Stata file into R.
The result is an R data.frame with attributes containing the metadata.
```{r}
library(readstata13)
cces_2018 = read.dta13("cces18_common_vv.dta", convert.factors = T, nonint.factors = T, generate.factors=T)
```

Fist we check the type, shape and contents of the sample.
```{r}
str(cces_2018, max.level=2, list.len=12, give.attr=FALSE)
dim(cces_2018)
names(cces_2018)[1:12]
names(attributes(cces_2018))
```

The `read.dta13` creates attributes corresponding to the Stata file metadata.
The attribute `names` corresponds to the dataframe column names and
the attribute `val.labels` provides a short description of that columns.
For questions which have categorical responses, the attribute `data.label`
provides the category name and interger code used for that survey question.
```{r}
names(attributes(cces_2018))
attr(cces_2018, "names")[1:10]
attr(cces_2018, "var.labels")[1:10]
attr(cces_2018, "label.table")[["gender"]]
```

The Common Content portion of the survey consists of questions asked of all subjects.
Profile questions provide demographic information,
and opinion questions provide outcomes of interest.
The demographic variables include
gender, education level, race, and geographic location, among others.
```{r}
table(cces_2018$gender[,drop=TRUE])
table(cces_2018$educ[,drop=TRUE])
table(cces_2018$race[,drop=TRUE])
```

The political opinion questions on the survey are labelled "President 2016", "Abortion", "Immigration", "Taxes", "Health Care".
```{r}
attr(cces_2018, "var.labels")[which(startsWith(attr(cces_2018, "var.labels"), "President 20"))]
pres_vote <- which(startsWith(attr(cces_2018, "var.labels"), "President 20"))
table(cces_2018[[pres_vote]][,drop=TRUE], useNA="always")

attr(cces_2018, "var.labels")[which(startsWith(attr(cces_2018, "var.labels"), "Health Care --"))[1]]
health_care_Q1 <- which(startsWith(attr(cces_2018, "var.labels"), "Health Care --"))[1]
table(cces_2018[[health_care_Q1]][,drop=TRUE], useNA="always")
```

Of the 525 columns in the survey, we are interested in the set of political opinion questons
and a subset of the demographics.  We create a new dataframe
which contains just these columns of interest.
Using the survey metadata, we select demographics by column names and political opinion questions by the column labels.
In order to properly postratify, we remove any record where there is missing demographic information.
In addition, so that we can easily align the CCES variables with the ACS variables
we removed unused category labels.

```{r}
col_labels <- attr(cces_2018, "var.labels")
col_names <- attr(cces_2018, "names")
demo_cols <- c(which(col_names == "birthyr"),
              which(col_names == "gender" ),
              which(col_names == "educ"),
              which(col_names == "race"),
              which(col_names == "hispanic"),
              which(col_names == "marstat"),
              which(col_names == "inputstate"))
# don't allow missing demographics
keep_rows <- complete.cases(cces_2018[, demo_cols])

outcome_cols <- c(which(startsWith(col_labels, "Abortion --")),
                 which(startsWith(col_labels, "Immigration --")),
                 which(startsWith(col_labels, "Taxes --")),
                 which(startsWith(col_labels, "Health Care --")),
                 which(startsWith(col_labels, "President 20")))
keep_cols <- sort(c(demo_cols, outcome_cols))

cces_2018_subset <- subset(cces_2018, keep_rows, keep_cols)
# `subset` drops attributes, keep attribute "var.labels"
attr(cces_2018_subset, "var.labels") <- attr(cces_2018, "var.labels")

# cleanup factor variables; drop unused category labels
for (name in names(cces_2018_subset)) {
    if (is.factor(cces_2018_subset[[name]]))  {
        cces_2018_subset[[name]] = droplevels(cces_2018_subset[[name]])
    }
}
```

@@

The CCES has 60,000 respondants.
According to the [Guide to the 2018 Cooperative Congressional Election Survey](https://sda.berkeley.edu/sdaweb/docs/cces2018/DOC/CCES+Guide+2018.pdf)

>This study constructed a very large sample capable of capturing variation across a wide variety of legislative constituencies. In fact, the state-level samples are sufficiently large as to measure with a reasonable degree of precision the distribution of voters’ preferences within most states.

While the CCES data may be roughly representative of the state-level political sentiment as a whole,
if we seek to understand certain subsets of the population,
it is unlikely to be representative of those groups.
There are over 2 million
possible combinations of gender, education,
race, marital status, birthyear, and state of residence.
If we bin ages by decade, there are a mere 200,000 combinations,
binning states by region further reduces this to 16,000, etc., etc.
However, if we wish to consider state-level subpopulations of interest,
the survey data is likely to be sparse and not representative of the general population.
To address the former we do multi-level modeling; for the latter, poststratification.

### Population data:  the 2018 American Community Survey

The US Census data is the largest representative survey of the US population that we have access to.
A full census is conducted every 10 years.
The American Community Survey (ACS) is a micro-census conducted every year by the Census Bureau
over 1% of the households in the US, i.e., it has 3 million respondants (as opposed to 300 million in the full census).

@@

The `ipumsr` package uses several [tidyverse](https://www.tidyverse.org/) packages
including [`dplyr`](https://dplyr.tidyverse.org/) and [`haven`](https://haven.tidyverse.org/).
The [`read_ipums_micro` function](http://tech.popdata.org/ipumsr/reference/read_ipums_micro.html)
returns a single [`tbl_df`](https://tibble.tidyverse.org/reference/tbl_df-class.html) data frame, i.e., a "tibble".
```{r}
library(labelled)
library(ipumsr)
load("acs_2018_subset.RData")
str(acs_2018_subset, max.level=2, list.len=6)
dim(acs_2018_subset)
names(acs_2018_subset)
```
We selected variables "RACE" and "HISPAN"; the extract contains two columns for each:
"RACE" and "RACED", and "HISPAN" and "HISPAND".  The "D" columns provide an extremely fine-grained
set of categories for race and hispanic origin.  As these go far beyond the categories from
the CCES which are provided, we ignore them.

The columns of this tibble are  [labelled](https://haven.tidyverse.org/reference/labelled.html) vectors.
Labelled vectors can have text labels associated with specific values.
To see how this works, we compare the values and labels for `AGE`.
The 2018 ACS data spans ages 0 through 96,
of these, only ages 0 and 90 have labels.
The function [`ipums_val_labels`](https://tech.popdata.org/ipumsr/reference/ipums_var_info.html)
allows us to inspect these labels.
```{r}
length(table(acs_2018_subset$AGE))
ipums_val_labels(acs_2018_subset$AGE)
ipums_val_labels(acs_2018_subset$AGE)
```

The function [`as_factor`](https://haven.tidyverse.org/reference/as_factor.html)
allows us to treat labelled vectors as factors.
See the vignette https://tech.popdata.org/ipumsr/reference/ipums_var_info.html for further information.
```{r}
head(table(acs_2018_subset$AGE))
head(table(as_factor(acs_2018_subset$AGE)))
```

### Aligning the survey and target population data with MRP-KIT

We use MRP-KIT to allow us to do inference on subpopulations of interest
using the following demographics:
sex, state of residence, education, race, marital status, and age.

MRP-KIT allows mappings that are one-to-one, many-to-one, one-to-many.
All mappings must be completely deterministic based on category label alone.
If this is not possible then you must create new, synthetic variables which can be
cleanly aligned.  Likewise, 
when information is spread across multiple columns in either the survey sample
or the target population sample, it must be combined into a single column's
worth of information.
If a demographic is recorded at too fine a level of granularity
and should be aggregated, e.g., binning date years into decades,
this much also be done beforehand.
MRP-KIT doesn't and cannot provide functions to do this; only you know
the data and what it means.
All such data pre-processing should be documented, srcipted, and *tested*
to insure replicability and/or use in production pipelines.

The MRP-KIT class `SurveyQuestion` is used to map the survey data demographic
information to the corresponding population data demographics.
The mapping process is both tedious and challenging.
The following table provides a summary of the demographic constructs
that we wish to use in analyzing the CCES dataset,
and challenges involved in mapping the CCES to the ACS.

Construct | CCES | ACS | Challenge
--------- | ---- | --- | ---------
Sex | gender | SEX | (none)
State | inputstate | STATEFIP | specify 1-1 mapping programatically
Education | educ | EDUC | different possible mappings
Race/Hispanic ethnicity | race, hispan | RACE, HISPAN | information spread across categories
Marital status | marstat | MARST | many-to-many mapping
Age | birthyr | AGE | different units, too fine-grained

#### Simple one-to-one mapping:  CCES "gender" / ACS "sex"

We create a new SurveyQuestion object with name "sex",
then create a mappng from the CCES to the ACS,
both for column names and response labels
where for both mappings, the first element
is the survey sample name/label, and the second element
is the population sample name/label.

@@

We examine the state names from both the CCES and ACS; the same labels are used in both surveys.
```{r}
# CCES data is a factor, levels are the labels
inputstate <- levels(cces_2018_subset$inputstate)

@@

We need to create a SurveyQuestion object, as above,  but it would be
extremely tedious to do this as a long list of tag, value pairs:
```
values_map=list("Alabama" = "Alabama", ..., "Wyoming" = "Wyoming")
```
The `values_map` is stored as an [R Pairlists](https://github.com/hadley/r-internals/blob/master/pairlists.md).
This is actually a labelled list, not a list of pairs, i.e.,
the expression `list("tag1" = "value1", "tag2" = "value2")`
creates a list of length 2 where each element is a single named item.
```{r}
foo <- list("tag1" = "value1", "tag2" = "value2")
foo
```
To create this list programmatically,
you must create a list of values and a list of labels,
then set the names of the values list to the labels.
```{r}
# create list, then set names (uppercase)
mapping <- as.list(STATEFIP)
names(mapping) <- toupper(inputstate)
mapping[1:3]

# the one-liner (lowercase)

mapping = as.list(setNames(STATEFIP, tolower(inputstate)))
mapping[1:3]
```

Having established the values map, we create the corresponding SurveyQuestion object.
We use accessor methods `name()`, `col_names()` and `values()` to check our work.
```{r}
mapping = as.list(setNames(STATEFIP, inputstate))
q_state <- SurveyQuestion$new(name = "state",
	col_names = c("inputstate", "STATEFIP"),
	values_map= mapping)
q_state$name()
q_state$col_names()
head(q_state$values())
```

#### Choice of mapping: education

The CCES and ACS samples record the education level of the respondant
in variables `educ` and `EDUC`, respectively.
Both the CCES and ACS record the highest level of education completed,
but the CCES records the degree obtained whereas the ACS records the year completed.
```{r}
names(table(cces_2018_subset$educ))
ipums_val_labels(acs_2018_subset$EDUC)
```
We choose to align these categories as follows.

CCES | ACS
---- | ---
"No HS" | "N/A or no schooling"
"No HS" | "Nursery school to grade 4"
"No HS" | "Grade 5, 6, 7, or 8"
"No HS" | "Grade 9"
"No HS" | "Grade 10"
"No HS" | "Grade 11"
"High School Graduate" | "Grade 12"
"Some College" | "1 year of college"
"2-year" | "2 years of college"
"2-year" | "3 years of college"
"4-year" | "4 years of college"
"Post-grad" | "5+ years of college"

Mpping "Some College" to both 1 year and 3 years of college is also possible,
reasoning that the goal of a 4-year college degree wasn't obtained.
The choice depends on both an understanding of the survey data and the particular goals
of your analysis.
MRP-KIT can only provide the mechanism for recording and carrying out the mapping.
```{r}
educ_mapping = list(
    "No HS" = "N/A or no schooling",
    "No HS" = "Nursery school to grade 4",
    "No HS" = "Grade 5, 6, 7, or 8",
    "No HS" = "Grade 9",
    "No HS" = "Grade 10",
    "No HS" = "Grade 11",
    "High School Graduate" = "Grade 12",
    "Some College" = "1 year of college",
    "2-year" = "2 years of college",
    "2-year" = "3 years of college",
    "4-year" = "4 years of college",
    "Post-grad" = "5+ years of college"
)
q_educ <- SurveyQuestion$new(name = "educ",
    col_names = c("educ", "EDUC"),
    values_map= educ_mapping)
q_educ$name()
q_educ$col_names()
q_educ$values()
```

#### Survey variable cannot be mapped to single census variable:  Race

The [IPUMS description of the ACS variable RACE](https://usa.ipums.org/usa-action/variables/RACE#description_section)
says:

>Currently, the Census Bureau and others consider race to be a sociopolitical construct,
not a scientific or anthropological one.  Many detailed RACE categories consist of national origin groups.

This is borne out by the CCES, where the question for variable "race" is
"What racial or ethnic group best describes you?",
and includes "Hispanic" as a possible response.
The ACS variable "RACE" has a different set of responses:
```{r}
names(table(cces_2018_subset$race))
ipums_val_labels(acs_2018_subset$RACE)
````

In order to use the ACS for poststratification, we need to create a variable
corresponding to the CCES variable "race" from a combination of ACS variables.
Both surveys contain variable "Hispanic Origin".
For the CCES, this is a "Yes/No" question:
"Are you of Spanish, Latino, or Hispanic origin or descent?".
In the ACS, the variable HISPAN identifies persons of Hispanic/Spanish/Latino origin and
classifies them according to their country of origin.
```{r}
names(table(cces_2018_subset$hispan))
ipums_val_labels(acs_2018_subset$HISPAN)
```

We propose to add a column to the ACS survey which combines information present in the `RACE` and `HISPAN` variables.
```{r}
# create column which blends "race" "hispan"
acs_race2 <- as_factor(acs_2018_subset$RACE)

# population percentages by label
prop.table(table(acs_race2))

# if "hispan" is "Yes", "race" label is "Hispanic".
is_hispan <- acs_2018_subset$HISPAN > 0 & acs_2018_subset$HISPAN < 5
length(is_hispan[is_hispan == TRUE]) / length(is_hispan)
levels(acs_race2) = c(levels(acs_race2), "Hispanic")
acs_race2[is_hispan==TRUE] <- "Hispanic"

# adjusted percentages
prop.table(table(acs_race2))

# add to acs_2018_subset data.frame - respect column types
acs_2018_subset$race2 = to_labelled(acs_race2)
```

A further possibility is to similarly combine the CCES variables "race" and "hispan" 
so that the question of "Hispanic origin" is "Yes", the race-hispan label is "Hispanic".
Were we to do this, we would create a new variable "race2" and add it to the
CCES subset so that we could evaluate both classifications.
```{r}
# Population percentages in CCES variable "race"
prop.table(table(cces_2018_subset$race))

# Intersection (and) vs. union (or) of race == "Hispanic" and hispanic == "Yes"
length(which(cces_2018_subset$race == "Hispanic" & cces_2018_subset$hispan == "Yes"))
length(which(cces_2018_subset$race == "Hispanic" | cces_2018_subset$hispan == "Yes"))

# create column which adjusts race labels
cces_race2 <- cces_2018_subset$race
cces_race2[cces_2018_subset$hispan == "Yes"] <- "Hispanic"
prop.table(table(cces_race2))

# add to acs_2018_subset data.frame - respect column types
cces_2018_subset$race2 = cces_race2
```

Once comparable variables are present in both surveys,
MRP-KIT can handle the mapping.
We choose to align CCES "race" with new ACS variable "race2" as follows

CCES "race" | ACS "race2"
----------- | -----------
"White" | "White"
"Black" | "Black/African American/Negro"
"Native American" | "American Indian or Alaska Native"
"Asian" | "Chinese"
"Asian" | "Japanese"
"Asian" | "Other Asian or Pacific Islander"
"Other" | "Other race, nec"
"Middle Eastern" | "Other race, nec"
"Mixed" | "Two major races"     
"Mixed" | "Three or more major races"
"Hispanic" | "Hispanic"

```{r}
race_mapping = list(
    "White" = "White",
    "Black" = "Black/African American/Negro",
    "Native American" = "American Indian or Alaska Native",
    "Asian" = "Chinese",
    "Asian" = "Japanese",
    "Asian" = "Other Asian or Pacific Islander",
    "Other" = "Other race, nec",
    "Middle Eastern" = "Other race, nec",
    "Mixed" = "Two major races",
    "Mixed" = "Three or more major races",
    "Hispanic" = "Hispanic"
)
q_race <- SurveyQuestion$new(name = "race",
    col_names = c("race", "race2"),
    values_map= race_mapping)
q_race$name()
q_race$col_names()
q_race$values()
```

This is yet another example of the inherent challenges of MRP.
The work of creating a mapping is an extension
of data assembly, validation, and exploratory data analysis,
all of which requires researcher domain knowledge.

#### Many-to-many mapping: marital status

The CCES variable "marstat" and ACS variable "MARST" record the marital status of the respondant.

```{r}
names(table(cces_2018_subset$marstat))
ipums_val_labels(acs_2018_subset$MARST)
```

For the 2018 CCES survey, the marital status of all respondants is known,
therefore categories "skipped" and "not asked" are not problematic.
The ACS variable ["MARST"](https://usa.ipums.org/usa-action/variables/MARST#comparability_section)
has two categories for corresponding to CCES category "Married":
"Married, spouse present", "Married, spouse absent".
Mapping both ACS categories to CCES category "Married" is a one-to-many mapping
which is allowed by MRP-KIT.
The CCES has category "Domestic / civil partnership",
for which there is no corresponding ACS category.
Mapping both partnerships and "Never married" to the ACS category "Never married/single"
is a many-to-one mapping from CCES to ACS, also allowed by MRP-KIT.
However, if we map CES category  "Domestic / civil partnership"
to either ACS category "Married, spouse present" or "Married, spouse absent",
we create a many-to-many mapping.

OK             |                | NOT OK         | 
-------------- | -------------- | -------------- | --------------
CCES | ACS | CCES | ACS
Married | Married, spouse present |Married | Married, spouse present 
Married | Married, spouse absent |Married | Married, spouse absent 
Separated | Separated |Separated | Separated 
Divorced | Divorced |Divorced | Divorced 
Widowed | Widowed |Widowed | Widowed 
Never married | Never married/single |Never married | Never married/single 
*Domestic / civil partnership* | *Never married/single* | *Domestic / civil partnership* | **Married, spouse present**

```{r, error=TRUE}
# This is OK
marstat_mapping = list(
    "Married" = "Married, spouse present",
    "Married" = "Married, spouse absent",
    "Separated" = "Separated",
    "Divorced" = "Divorced",
    "Widowed" = "Widowed",
    "Never married" = "Never married",
    "Domestic / civil partnership" = "Never married"
)
q_marstat <- SurveyQuestion$new(name = "marstat",
    col_names = c("marstat", "MARST"),
    values_map= marstat_mapping)
q_marstat$name()
q_marstat$col_names()
q_marstat$values()

# This is NOT
marstat_mapping = list(
    "Married" = "Married, spouse present",
    "Married" = "Married, spouse absent",
    "Separated" = "Separated",
    "Divorced" = "Divorced",
    "Widowed" = "Widowed",
    "Never married" = "Never married",
    "Domestic / civil partnership" = "Married, spouse present"
)
q_marstat <- SurveyQuestion$new(name = "marstat",
    col_names = c("marstat", "MARST"),
    values_map= marstat_mapping)

```
Here we propose to add yet another column to the ACS which conflates "Married, spouse present" and "Married, spouse absent".
